A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-09 tex/BiSig.tex
[ tex ] Revise S3
[ tex ] Minor changes
[ tex ] Merging from the old master branch
[ tex ] Minor in S3; adjust the todonote configuration
[ tex ] Fix typos and revise S5

\newcommand{\LT}[1]{\todo[author=LT,inline,color=orange!40]{{#1}}}
\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=yellow!40]{{#1}}}

@@

\begin{abstract}
%  Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
  While there is much progress in designing bidirectional type systems and associated type synthesis algorithms on a case-by-case basis, the lack of a general theory makes it difficult to study the common principles and develop widely applicable techniques and practical tools such as a `type-synthesiser generator'.
  In response to this difficulty, we propose a simple yet general and constructive theory of bidirectional type synthesis.
%  which refers to the mutual use of type checking and synthesis based on bidirectional typing rules to determine the type of a program with type annotations.
  First, it is simple yet general as it applies to any syntax-directed bidirectional simple type system that can be specified by a signature.
%  which is analogous to a grammar in parsing.
  Second, it is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also compactly formulated in a way that unifies computation and proof---for example, a proof of the decidability of the type synthesis problem is computationally a type synthesiser.
  Moreover, we introduce `mode preprocessing' to deal with missing type annotations, an aspect often overlooked in the literature.
  The main theorem is logically a trichotomy on untyped abstract syntax trees and computationally a type synthesiser that may throw an exception due to missing type annotations.
  By being parametrised by a signature, our theory works as a verified type-synthesiser generator overall.
\end{abstract}

@@

\ccsdesc[500]{Software and its engineering~Compilers}
\ccsdesc[500]{Software and its engineering~Syntax}
\ccsdesc[500]{Theory of computation~Type theory}
\ccsdesc[300]{Software and its engineering~Formal software verification}

\keywords{type checking, type inference, language formalisation, \Agda}

@@

\begin{acks}
%We thank Nathanael Arkor for the useful conversation.
The work is supported by the Ministry of Science and Technology of Taiwan under grant MOST 109-2222-E-001-002-MY3.
\end{acks}

>>>>

\newcommand{\LT}[2][]{\todo[author=LT,color=orange!40,#1]{#2}}
\newcommand{\Josh}[2][]{\todo[author=Josh,color=yellow!40,#1]{#2}}

@@

\begin{abstract}
  \LT{Still not coherent enough}
%  Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
  While there is much progress in designing bidirectional type systems and associated type synthesis algorithms on a case-by-case basis, the lack of a general theory makes it difficult to study the common principles and develop widely applicable techniques and practical tools such as a `type-synthesiser generator'.
  In response to this difficulty, we propose a simple yet general and constructive theory of bidirectional type synthesis.
%  which refers to the mutual use of type checking and synthesis based on bidirectional typing rules to determine the type of a program with type annotations.
  First, it is simple yet general as it applies to any syntax-directed bidirectional simple type system that can be specified by a signature.
%  which is analogous to a grammar in parsing.
  Second, it is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also compactly formulated in a way that unifies computation and proof---for example, a proof of the decidability of the type synthesis problem is computationally a type synthesiser.
  Moreover, we introduce `mode preprocessing' to deal with missing type annotations, an aspect often overlooked in the literature.
  The main theorem is logically a trichotomy on untyped abstract syntax trees and computationally a type synthesiser that may throw an exception due to missing type annotations.
  By being parametrised by a signature, our theory works overall as a verified type-synthesiser generator.
\end{abstract}

@@

\ccsdesc[500]{Software and its engineering~Compilers}
\ccsdesc[500]{Software and its engineering~Syntax}
\ccsdesc[300]{Software and its engineering~Formal software verification}
\ccsdesc[300]{Theory of computation~Type theory}

\keywords{datatype-generic programming, language formalisation, \Agda}

@@

\begin{acks}
%We thank Nathanael Arkor for the useful conversation.
We thank Kuen-Bang {Hou (Favonia)} for his comments and suggestions.
The work is supported by the Ministry of Science and Technology of Taiwan under grant MOST 109-2222-E-001-002-MY3.
\end{acks}
